---
title: "Download and Process Independent Variables"
author: "steppe"
date: "2024-01-12"
output: html_document
---

We have a named list of variables which we are interested in. 
We will download each of these variables using curl. 
```{sh download Chelsa climate variables, eval = F}
URL_LIST=$(<../data/raw/envidatS3paths.txt)
## echo $URL_LIST | xargs -n 1 -P 4 wget -q -P ../data/spatial/
```

```{r}
library(terra)
library(sf)
```

```{r define domain for cropping}
domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
crs(domain) <- "EPSG:4326"
```

Subset the CHELSA variables to domain of analysis. 
```{r subset CHELSA, eval  = F}

p <- '../data/spatial/raw'
chelsa <- rast(file.path(p, list.files(p)))
domain <- project(domain, crs(chelsa))
chelsa <- crop(chelsa, domain)

terra::writeRaster(chelsa, "../data/spatial/processed/CHELSA.tif")

# sudo du -sh /raw ... 2.7 gb
# sudo du -sh processed  ... 214 MB 
# big space saver!

rm(p, chelsa)
```

I missed the download of the last GDD! We ran that with a slight modification to the bash code above, changing the URL file, and process these variables now. 

```{r chelsa2, eval = F}

p <- '../data/spatial/raw'
chelsa <- rast(file.path(p, list.files(p)))
domain <- project(domain, crs(chelsa))
chelsa <- crop(chelsa, domain)
chelsa <- subst(chelsa, NA, 0)
names(chelsa) <- names(chelsa) <- gsub('CHELSA_|_1981.*$', '', names(chelsa))

chelsa_og <- terra::rast( "../data/spatial/processed/CHELSA.tif")
chelsa <- c(chelsa_og, chelsa)
terra::writeRaster(chelsa, "../data/spatial/processed/CHELSA1.tif", overwrite = TRUE)
```

We will also download soilgrids 250m data for bulk density. Bulk density is defined by soil texture, organic carbon, and the packing of particles. Soils with higher bulk density have been identified as holding less soil moisture content. 

```{r download soilgrids data, eval = F}

library(gdalUtilities)

chelsa <- rast("../data/spatial/processed/CHELSA.tif")
domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
ext(domain) <- c( -125.5, -100, 25,  50) # set the extent
crs(domain) <- "EPSG:4326"
igh='+proj=igh +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs'
bb1 <- as.polygons(domain) |>  # convert to vector data
  st_as_sf() |> # convert to simple feature
  st_transform(igh) %>% 
  st_bbox()

bb <- c(bb1$xmin, bb1$ymax, bb1$xmax, bb1$ymin)
sg_url = "/vsicurl?max_retry=3&retry_delay=1&list_dir=no&url=https://files.isric.org/soilgrids/latest/data/"

gdal_translate(paste0(sg_url,'bdod/bdod_15-30cm_mean.vrt'),
    "../data/spatial/crop_roi_igh_r.tif",
    of="VRT",tr=c(250,250),
    projwin=bb,
    projwin_srs = igh)

bdod <- rast('../data/spatial/crop_roi_igh_r.tif')
bdod <- project(bdod, chelsa, align = TRUE, threads = 20, 
                filename = '../data/spatial/processed/bulk_density.tif', overwrite = T)

bdod <- crop(bdod, chelsa)
names(bdod) <- 'bulk_density'
writeRaster(bdod, filename = '../data/spatial/processed/bulk_density.tif', overwrite = T)

rm(igh, bb1, bb, sg_url, bdod)
```

```{r process cti, eval = F}

cti <- rast('~/Downloads/dtm_cti_merit.dem_m_250m_s0..0cm_2018_v1.0.tif') # 6gb big pup. 
cti <- crop(cti, domain)
cti <- resample(cti, chelsa, 
                filename = '../data/spatial/processed/cti.tif', threads = 20)
cti <- rast('../data/spatial/processed/cti.tif')
names(cti) <- 'cti'
writeRaster(cti, '../data/spatial/processed/cti.tif', overwrite = TRUE)

rm(cti)
```

the growing degrees data use NA instead of 0 or one for areas without freezing temps
```{r rename layers and replace NA, eval = F}
che <- rast('../data/spatial/processed/CHELSA.tif')
che <- subst(che, NA, 0)
names(che) <- gsub('CHELSA_|_1981.*$', '', names(che))
writeRaster(che, '../data/spatial/processed/CHELSA.tif', overwrite = TRUE)
```

Mask the data to terrestrial surfaces
```{r}
che <- rast('../data/spatial/processed/CHELSA.tif')
ocean <- rnaturalearth::ne_download(category = 'physical', type = 'ocean', scale = 10)
che <- mask(che, ocean, inverse= T)

writeRaster(che, '../data/spatial/processed/CHELSA.tif', overwrite = TRUE)

rm(ocean)
```



create a raster of latitude, this will serve as our surrogate for day length

```{r}
che <- rast('../data/spatial/processed/CHELSA.tif')
lat <- init(che, 'y')

writeRaster(che, '../data/spatial/processed/latitude.tif', overwrite = TRUE)
```

