--- 
title: 'Phenology'
author:  |
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org]
abstract: |
  \noindent
  **Premise:** word to the birds    
  **Methods and Results:** birds on the wings  
  **Conclusions:** birds are not real     
keywords: |
  phenology, herbarium, gamm, native seed   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
--- 

```{r echo = F}
knitr::opts_chunk$set(echo=F, warning = F, message = F)
```

# Introduction

Changes in a species phenology, the timing of life history events, are one of the most common and pronounced responses to climate change [@parmesan2003globally]. 
Accordingly, considerable effort has been directed towards exploring the causal links between climate change and phenology [@tang2016emerging]. 
Given the importance of phenology to biodiversity, and the ready identification of causal agents, many meteorological explanatory variables have been produced, as well as remotely sensed vegetation attributes linked to phenology, e.g. vegetation stand wide leaf out, and leaf senescence dates [@dronova2022remote]. 
Generally studies which tend to treat species as vegetation complexes or communities have found that early season phenophases, any noticeable stage in a life cycle, have tended to advance, while late season phenophases have often become delayed [@parmesan2003globally]. 
However, for the individual plant species analyzed to date, responses to warming have been idiosyncratic preventing the generalization of results across, clades, most functional groups, resulting in a need for continued species specific modelling of phenology for many applications [@caradonna2014shifts; @augspurger2020concordance]. 
These observational studies and manipulative experiments have generally been limited to a few dozen species in only one to a couple populations, or when using herbarium sheets from many populations across a spatial domain - than only a few species [@katal2022deep], although some exceptions exist s[@park2023complex].

Rather than the documentation of trends over time, e.g. flowering initiation advancing by 2.3 days per decade, associated with phenology research in a climate change context the capability to predict the timing of phenophases in an individual year based on realized weather are required for several applications, most notably agriculture and related disciplines. 
Species specific models have been generated for crop varieties for over half a century [@hodges1990predicting], and increasingly incorporate data sources which seldom exist for wild species, e.g. genes, near real-time remote sensing data of pure stands of individuals, and large amounts of training data capable of training artificial intelligence [@deva2024new; @nagai2020simpler; @gao2021mapping]. 
Additionally, these agricultural systems minimize several environmental factors e.g. the severity of drought, and are operating on lineages breed for consistent windows of phenophases; hence these recent innovations in crop science are difficult to transfer to wildland settings. 

The number of papers attempting to predict flowering events in an individual year, across geographically large portions of a species range are fewer than either of the above use-cases [@hodgson2011predicting]. 
Attempting to model these events are complicated because not only are the responses of individual species idiosyncratic to climate change, the response of populations varies across species ranges, due not only to differing levels of climate change, but to existing broad environmental climate [@park2019herbarium; @park2023complex] 
Generalized Additive Mixed Models (GAMM's) are often used to document a phenophase because a single model can have their splines fit to both initiation, peak, and cessation of an phase, using a single or multiple independent variables - a limitation of several other methods of estimation [@polansky2013generalized].
The use of independent variable(s) alongside GAMM's ability to incorporate an error-correlation structure which accommodates spatial autocorrelation allows them to model the phenological parameters of a species across it's geographic and concomitant environmental range. 
However, the data sets which cover the wide range of species which may be desired to model are few, with herbaria and citizen science initiatives being the two largest sources. 


Disciplines which straddle ecology and agriculture, such as wildland seed harvest, require useful models of major phenological events (flowering, and fruit dispersal) to optimize the detection of populations, estimates of census sizes, and the eventual collection of native seed.  
Hererin we use GAMM's to model phenophases, inferred from herbarium specimens and using environmental predictors identified as casual cues of phenology, in space. 
Our necessity to more accurately understand the phenology of species arose from our goal of native seed collection for both native plant germplasm development, and *ex-situ* conservation. 
The identification of putative populations with enough individuals to warrant germplasm development is a time consuming process, because most plant species can only be identified when they have reproductive organs, and a populations ability to support these collections varies wildly with the years weather, pathogen load, and various stochastic processes. 
The collection of seeds, which is generally occurring for both many species and many populations each year, is challenged by both the need for crews to collect from other species and the natural dispersal of seeds - simply put *timing is everything*.  

# Methods

## Data Sources & Cleaning

Species records were derived from the Symbiota herbarium portal for all years from 1981-2021, these years reflected the climate means used as independent variables (ridigbio). 
All records were downloaded, and the records in the 2.5% Day of Year (DOY) quantile were manually reviewed. 
These early records were reviewed because novice collectors, especially with graminoids, may actually collect material without reproductive organs yet reaching anthesis ('in bud').
The later records were reviewed because collectors may have collections of individuals entirely post-anthesis - a situation very common with certain clades where species are commonly distinguished by morphological characteristics of their fruits (e.g. the Fabaceae or Leguminosae). 
In both scenarios analysts proceeded towards the mean of the distribution until they encountered 5 consecutive sheets with the desired phenophase.  

| Layer |                       Description                       |              Source                   |     Abbrev   |      
| :---: | :-----------------------------------------------------: | :-----------------------------------: | :----------: | 
|  1.   |        Mean Temperature of Warmest Quarter (BIO10)      |              Chelsa                   |     bio10    |
|  2.   |             Precipitation of Driest Month (BIO14)       |              Chelsa                   |     bio14    |
|  3.   |      Mean Monthly vapour pressure deficit (vpd)         |              Chelsa                   |   vpd_mean   |
|  4.   |     Heat accumulation of  Degree-days above 0C (gdd0)   |              Chelsa                   |     gdd5     |
|  5.   |         First growing degree day above 0C (gdgfgd0)     |              Chelsa                   |    gdgfgd5   |
|  6.   |           Number of Degree-days above 0C (ngd0)         |              Chelsa                   |     ngd5     |
|  7.   |     Heat accumulation of  Degree-days above 5C (gdd5)   |              Chelsa                   |     gdd5     |
|  8.   |         First growing degree day above 5C (gdgfgd5)     |              Chelsa                   |    gdgfgd5   |
|  9.   |           Number of Degree-days above 5C (ngd5)         |              Chelsa                   |     ngd5     |
| 10.   |   Heat accumulation of  Degree-days above 10C (gdd10)   |              Chelsa                   |     gdd10    |
| 11.   |         First growing degree day above 10C (gdgfgd10)   |              Chelsa                   |   gdgfgd10   |
| 12.   |         Number of Degree-days above 10C (ngd10)         |              Chelsa                   |     ngd10    |
| 13.   |             Net Primary Productivity (npp)              |              Chelsa                   |      npp     |
| 14.   |                       Soil Bulk Density                 |             SoilGrids                 |      sbd     |
| 15.   |                  Compound Topographic Index             |               terra                   |      cti     |


Independent variables reflected climate, and landform and soil parameters which modulate soil moisture. 
The climate variables from CHELSA, were 1981-2010 annual means, for Growing Degree Days (GDD) heat sums (at 0°C, 5°C, 10°C), first (gdgfgd) and last (gddlgd) GDD DOY, vapor pressure deficit (vpd), Bio10 (mean daily mean air temperatures of the warmest quarter), and Bio14 (precipitation amount of the driest month). 
Soil bulk density, which is shown to reflect the amount of air/water space in soil, was downloaded from SoilGrids. 
Compound Topographic Index (cti), which describes the potential of an area to accumulate soil moisture via a combination of its landform position, slope, aspect, and size of it's up-slope catchment area, was downloaded from geomorpho90m and resampled from 90m to the 250m resolution of the previous data sets. 

GAM's require data on when a species was **not** flowering in order to develop splines for the onset of flowering. 
Pseudo-floral absences were created using known sites, and their observed phenology. 
All of the CHELSA climate variables were decomposed using PCA, and the first axis (explaining 98.1% of the variation; 750m x 750m cells) was used as a feature space in a Ward-like hierarchical clustering algorithm which seeks to maximize homogeneity of both the feature and constraint space - here geography. 
A suitable number of clusters from the independent variable were automatically selected using kgs [@white2022maptree], these clusters were then reanalyzed in light of the constraint space using automatic selection of an alpha parameter which blends the feature and constraint space and re-clustered using hclustgeo [@chavent201clustgeo]. 
Each cluster had weibull estimates of flowering initiation and cessation modelled, and any DOY within 28 days preceding onset or following cessation were drawn for each group [@belitz2020phenesse]. 
These values were arranged by ascending DOY and joined to the members of the group via decreasing warm to cool values along the PCA axis. 
Points in clusters which had a nearest geographic neighbor in another cluster had their randomly generated pseudo-absences wiped, and thin plate spline regression using the PCA axis as an independent variable, and interpolation was used to repopulate the floral pseudo-absence dates [@nychka2021fields, @hijman2024terra]. 

## Modelling

All independent variables were extracted to the dependent variables, and if a value for an independent variable was missing - which was not uncommon for Soil Bulk Density, where the modellers excluded the fringes of several vernally wet playas - it was imputed as the mean of the variable for the species. 
All independent variables then underwent feature selection using the Recursive Feature Elimination (rfe) with 10 Cross-Validations (CV) folds, 5 replicates, and from 1-10 variables using caret [@kuhn2008caret].
The remaining variable(s) were used as covariates with DOY always included in the models. 
A GAMM was fit using presence/absence of flowering as a response, as well as GAMM's with error structure of gaussian, spherical, and exponential variograms, with REML [@barton2023mumin]. 
All models were subjected to model selection, and the top model determined via AUC scores [@barton2023mumin]. 

## Surfaces

Each species top model was used to determine start and end dates (DOY) for which flowering was likely to occur, while initially omitting space as an explicit variable.   
These predicted start and end dates were used to constrain the number of models predicted onto raster surfaces at each time point, a computationally intensive process. 
This model was fit using an initial prediction matrix spanning the range of all independent variables at 15 points across each of their ranges, based on minima and maxima observed in the dependent data. 
The first DOY with a > 55% probability of flowering was used as the start date, and the lat DOY with a >60% probability of flowering, were used as constraints for spatial modelling. 
The higher tolerance for flowering cessation was used because the distribution of flowering generally follows a LONG TAIL distribution. 

Models were predicted onto rasters at biweekly (14 day) intervals from the start to end DOY, in areas which Species Distribution Models predicted as having a high probability (> 60%???) of suitable habitat (Benkendorf et al. 202X). 
Rasters which had fewer than 5% of their total cells classified as having a >50% probability of flowering were subsequently discarded.

## Interpretation 

# Results 
```{r}
library(tidyverse)
```


## Herbarium Records 

```{r}
scored <- bind_rows( 
  read.csv('../data/processed/high_priority_sheets-scored.csv'), 
  read.csv('../data/processed/second_phenology_review-scored.csv'), 
  read.csv('../data/processed/low_priority_sheets-scored.csv'), 
  read.csv('../data/processed/troublesomeSpecies-scored.csv') 
  ) %>% 
  mutate( 
    Anthesis = if_else(is.na(Anthesis), 999, Anthesis) 
    ) %>% 
  select(-comments) 

PrcntFlr <- scored %>% 
  group_by(scntfcnm) %>% 
  filter(Anthesis != 999) %>% 
  summarise(PrcntFlr = sum(Anthesis)/n()) 

# write.csv(PrcntFlr, '../data/processed/PrcntFlr.csv', row.names = F)
PrcntFlr <- read.csv('../data/processed/PrcntFlr-modded.csv') |>
  filter(!Lifeform %in% c('Liana', 'Succulent'))
```


```{r do life forms vary, results='hide'}
res_aov <- aov(PrcntFlr ~ Lifeform, data = PrcntFlr)

# car::qqPlot(res_aov$residuals) # deviates way outside. 
# shapiro.test(res_aov$residuals) # reject null hypothesis data are drawn from a normal distribution. 

kw <- kruskal.test(PrcntFlr ~ Lifeform, data = PrcntFlr)
```

```{r boxplot lifeform percent anthesis, eval=FALSE}
# pairwise.wilcox.test(PrcntFlr$PrcntFlr, PrcntFlr$Lifeform, p.adjust.method = "BH")

ggpubr::ggboxplot(PrcntFlr, x = "Lifeform", y = "PrcntFlr", 
          color = "Lifeform", #palette = c("#00AFBB", "#E7B800", "#FC4E07"),
          order = c("Forb", "Graminoid", "Shrub", 'Tree'),
          ylab = "Proportion Flowering", xlab = "Lifeform") + 
  #ggpubr::stat_compare_means() + 
  ggpubr::geom_pwc(
    method = "t_test", label = "p.adj.format",
    p.adjust.method = "BH", hide.ns = TRUE,
    bracket.nudge.y = 0.2
  ) +
  geom_jitter(alpha = 0.2, width = 0.2)
```

A total of `r length(unique(scored$scntfcnm))` species had a total of `r nrow(scored)` herbarium sheets reviewed. 
Of these sheets `r sum(scored$Anthesis == 0)` (`r round(sum(scored$Anthesis == 0)/nrow(scored) *100, 1)`%) were not in anthesis. 
`r nrow(PrcntFlr[PrcntFlr$PrcntFlr <= 0.50,])` of these species had less than 50% of their records in anthesis.  
There was strong evidence that lifeforms varied in the proportion of mean number of sheets which were flowering (kruskall wallis p = `r kw$p.value`), with more forbs flowering than each of the three other life forms, and some evidence that more graminoids than trees being in flower. 

## Modelling 

```{r Variation in initiation and cessation dates}

p <- '../results/initation_cessation_tables'
clustered_initCess <- do.call(rbind,
          lapply(file.path(p, list.files(p)), readr::read_csv, id = "taxon", 
                 show_col_types = FALSE)) |>
  mutate(taxon = gsub('[.]csv', '', basename(taxon))) |>
  group_by(taxon)

clusts <- slice_max(clustered_initCess, ClusterID) |>
  pull(ClusterID) # Number of clusters identified per species. 

clustered_init <- clustered_initCess |> 
  filter(event == 'initiation') |>
  arrange(estimate, .by_group = T) |>
  mutate(
    est_diff = estimate - lag(estimate),
    low_diff = low_ci - lag(low_ci),
    high_diff = high_ci - lag(high_ci)
  )

clustered_cess <- clustered_initCess |> 
  filter(event == 'cessation') |>
  arrange(estimate, .by_group = T) |>
  mutate(
    est_diff = estimate - lag(estimate),
    low_diff = low_ci - lag(low_ci),
    high_diff = high_ci - lag(high_ci)
  )

ci <- clustered_init %>%
 slice(c(1, n())) |>
  mutate(
    est_diff = estimate - lag(estimate),
    low_diff = low_ci - lag(low_ci),
    high_diff = high_ci - lag(high_ci)
  ) 

cc <- clustered_cess %>%
 slice(c(1, n()))|>
  mutate(
    est_diff = estimate - lag(estimate),
    low_diff = low_ci - lag(low_ci),
    high_diff = high_ci - lag(high_ci)
  ) 
```

Variation in Initiation and Cessation Dates between Groups

The number of clusters recovered by hclustgeo ranged from `r min(clusts)` to `r max(clusts)` with a median of `r median(clusts)`. 
The difference in the estimated day of year which flowering started between each of these clusters within a species was mean = `r mean(clustered_init$est_diff, na.rm = T)`, median = `r median(clustered_init$est_diff, na.rm = T)`, and the difference in the estimated day of year flowering ended was mean = `r mean(clustered_cess$est_diff, na.rm = T)`, median = `r  median(clustered_cess$est_diff, na.rm = T)`.
The mean difference between the earliest and latest initiation estimate were mean = `r mean(ci$est_diff, na.rm = T)` median = `r median(ci$est_diff, na.rm = T)`, and for cessation events was mean = `r mean(cc$est_diff, na.rm = T)`, median = `r median(cc$est_diff, na.rm = T)`. 

```{r Converged models and terms}

covariate_extractor <- function(x){
  tmp <- unlist(strsplit(as.character(x$terms[[3]][-1]), split = ' [+] '))
  tmp <- tmp[grep('doy', tmp, invert = T)]
  tmp <- data.frame(t(tmp))
  colnames(tmp) <- paste0('covariate', 1:ncol(tmp))
  return(tmp)
}

covariates <- lapply(file.path('../results/models', list.files('../results/models')), readRDS)
covariates <- lapply(covariates, covariate_extractor) |>
  bind_rows() |>
  mutate(Taxon = gsub('[.]rds', '', list.files('../results/models')), .before = 1) |>
  pivot_longer(cols = starts_with('covariate'), names_to = 'Covariate', values_to = 'Variable') |>
  drop_na(Variable)

covariates %>% 
  group_by(Taxon) %>% 
  mutate(Variable_Relate = lag(Variable))

ggplot() + 
  geom_bar(data = covariates, aes(x = Variable))

```

How many models converged? 
Which variables ended up making it to the top models? Is this different than species mean LATITUDE??? 

## Ground Verification 

What percent of scouting records phen for flowering were within bounds? 
When was peak % flowering observed? 

# Discussion

```{r}
```

