---
title: "Species to model Phenology"
author: "steppe"
date: "2024-01-10"
output: html_document
---

```{r}
library(tidyverse)
library(sf)
library(terra)
source('functions.R')
set.seed(42)
```

```{r import species}

spp <- read.csv('../data/raw/2024species-Priorities.csv') |>
  unite('scientific_name',  genus:infraspecies, na.rm = TRUE, sep = ' ', remove = F) |>
  mutate(
    scientific_name = trimws(scientific_name)) |>
  select(scientific_name, genus, species, contract, priority, priority_source) |>
  count(scientific_name, priority)

```

```{r}
high_p <- spp %>% 
  filter(priority == 'High') %>% 
  arrange(-n)

low_p <- spp %>% 
  filter(priority != 'High' & ! scientific_name %in% high_p$scientific_name )
```

```{r pull sheets for species and subset to original, eval = F}

specs1 <- lapply(spp$scientific_name, specimen_sampler, 
                 date = '1981-01-01')

specs1 <- specs1[ lapply(specs1, typeof) == 'list' ] |> bind_rows()
# specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp', quiet = TRUE)
specs1 <- mutate(specs1, scntfcn = str_to_sentence(scientificname)) %>% 
  filter(scientificname != 'Chaenactis artemisifolia')
specs1 <- split(specs1, f = specs1$scntfcn)
specs1 <- lapply(specs1, rec_selec)
specs1 <- bind_rows(specs1)

dir.create('../data/processed/high_priority_sheets')
sf::st_write(specs1, '../data/processed/high_priority_sheets/high_priority_sheets.shp', 
             quiet = TRUE, append = F)

rm(specs1)
```

```{r flag records at outer quantiles of flowering, eval = F}

# extract pca values
preds <- rast('../results/spatial/gddPCA.tif')
specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp')

specs1 <- terra::extract(preds, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

# flag records for manual review. 
specs1 <- split(specs1, f = specs1$scntfcn)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.025, 0.975))
specs2 <- bind_rows(specs1)

priority_spp <- spp %>% 
  filter(priority %in% c('Medium', 'High') | n > 1) %>% 
  pull(scientific_name)

specs2 %>%
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% 
  filter(scntfcnm %in% priority_spp) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "") #%>% 
 # write.csv( '../data/processed/high_priority_sheets.csv', row.names = F, )


rm(preds, specs2, specs1)

as.Date(206, origin = '1987-01-01')
```

the above records were flagged by visual review of the specimen. During this process, the realization that the check for bimodality should have been performed, and the verification that certain species have many more records outside anthesis, called for a second batch of scoring.

```{r detect multimodal data, eval = F}

d <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp') %>% 
  mutate(scntfcnm = str_to_sentence(scntfcnm))
splicies <- split(d, f = d$scntfcnm)
ou <- sapply(X = splicies, FUN = function(x){diptest::dip(x$doy) }) # this test seems useless. 
ou <- ou[ou < 0.03]
sort(ou, decreasing = F)

ob <- lapply(splicies, modal_finder, path = '../results/density_curves2/')
ob <- data.frame(scntfcnm = names(ob), doy = as.numeric(ob))
ob1 <- mutate(ob, doy = case_when(
  scntfcnm == 'Acamptopappus sphaerocephalus' ~ 215,
  scntfcnm == 'Achnatherum speciosum' ~ 250,
  scntfcnm == 'Adenostoma sparsifolium' ~ 999,
  scntfcnm == 'Astragalus eremiticus' ~ 240,
  scntfcnm == 'Atriplex canescens' ~ 999,
  scntfcnm == 'Atriplex confertifolia' ~ 999,
  scntfcnm == 'Balsamorhiza serrata' ~ 180,
  scntfcnm == 'Centromadia elegans' ~ 999,
  scntfcnm == 'Cylindropuntia bigelovii' ~ 210,
  scntfcnm == 'Cylindropuntia bigelovii' ~ 210,
  scntfcnm == 'Danthonia californica' ~ 999,
  scntfcnm == 'Erigeron breweri' ~ 999,
  scntfcnm == 'Lomatium nudicaule' ~ 999,
  scntfcnm == 'Machaeranthera canescens' ~ 999,
  scntfcnm == 'Machaeranthera gracilis' ~ 190,
  scntfcnm == 'Muhlenbegia rigens' ~ 999,
  scntfcnm == 'Penstemon acuminatus' ~ 999,
  scntfcnm == 'Phlox stansburyi' ~ 210,
  scntfcnm == 'Psorothamnus schottii' ~ 225,
  scntfcnm == 'Rosa californica' ~ 999,
  scntfcnm == 'Rubus ursinus' ~ 999,
  scntfcnm == 'Sambucus mexicana' ~ 999,
  scntfcnm == 'Sphaeralcea munroana' ~ 999,
  scntfcnm == 'Stanleya pinnata' ~ 999,
  scntfcnm == 'Symphyotrichum subulatum' ~ 999,
  scntfcnm == 'Triteleia hyacinthina' ~ 999,
  .default = as.numeric(doy)
)) %>% 
  filter(doy != 999)


rm(ou, splicies, ob)
```

After a single round of scoring quantile values, we will identify whether certain species have many collections which are made outside of anthesis.

```{r identify collections commonly made outside of anthesis, eval = F}

d <- read.csv('../data/processed/high_priority_sheets-scored.csv') %>% 
  mutate(comments = na_if(comments, '')) %>% 
  filter((comments != "str") %>% replace_na(TRUE)) %>% 
  drop_na(Anthesis) 

d1 <- d %>% # keep records which weren't wonky. 
  group_by(scntfcnm, Anthesis) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs)

filter(d1, n() == 1) # totally clean
filter(d1, n() == 2) %>% 
  filter(Anthesis == 0, nAnt <= 0.5)
dirty <- filter(d1, n() == 2) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)

rm(d1)
# for the dirty species, we will determine whether both sides have specimens
# outside the desired phenophase or whether only  a single side does. 
# we will only evaluate specimens on the afflicted end(s). 

dirty1 <- filter(d, scntfcnm %in% dirty$scntfcnm) %>% 
  group_by(scntfcnm) %>% 
  mutate(
    avg_fl = mean(doy), 
    period = if_else(doy < avg_fl, 'E', 'L')
    ) %>% 
  group_by(scntfcnm, Anthesis, period) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs
  ) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)
  
rm(dirty, d)
```

Identify records in quantiles.

```{r identify records for second batch of scoring, eval = F}

# extract pca values
preds <- terra::rast('../results/spatial/gddPCA.tif')
specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp',
                  quiet = T) %>% 
  filter(scntfcnm %in% c(dirty1$scntfcnm, ob1$scntfcnm))

specs1 <- terra::extract(preds, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

rm(preds)
# split records with bimodal flowering. 
bimodals <- inner_join(specs1, rename(ob1, 'split' = doy), by = 'scntfcnm') %>% 
  mutate(split = if_else(doy < split, 'EARLY', 'LATE'))

split_bimodals <- split(bimodals, f = ~ bimodals$scntfcnm + bimodals$split)
split_bimodals <- lapply(split_bimodals, visual_review_flagger, probs = c(0.025, 0.975))

sb <- split_bimodals %>%
  map(mutate, phen_flag = as.character(phen_flag)) %>% 
  bind_rows() %>% 
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% # remove records which we already scored, and restrict to 
  # our species of interest.
  filter(scntfcnm %in% ob1$scntfcnm, !accessr %in% d$accessr) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "")

# flag records for further manual review. 
specs1 <- split(specs1, f = specs1$scntfcnm)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.05, 0.95))

dirty1 <- ungroup(dirty1)
specs2 <- specs1 %>%
  bind_rows() %>% 
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% # remove records which we already scored, and restrict to 
  # our species of interest.
  filter(scntfcnm %in% dirty1$scntfcnm, !accessr %in% d$accessr) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "", # select only the troublesome time period
         period = if_else(doy < mean(doy), 'E', 'L')) %>% 
  inner_join(., select(dirty1, scntfcnm, period), by = join_by(scntfcnm, period)) %>% 
  select(-period) %>% 
  filter(! accessr %in% sb$accessr)

bind_rows(sb, specs2) %>% 
  arrange(scntfcnm, doy) %>% 
  write.csv(., '../data/processed/second_phenology_review.csv', row.names = F)

rm(sb, specs2, dirty1, split_bimodals, bimodals, ob1, specs1)
```

Perform scoring on species with lower levels of importance

```{r}
spps_scored <- read.csv('../data/processed/high_priority_sheets-scored.csv') %>% 
  distinct(scntfcnm) %>% 
  pull(scntfcnm)

specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp',
                  quiet = T) %>% 
  filter(! scntfcnm %in% spps_scored)

# extract pca values
preds <- rast('../results/spatial/gddPCA.tif')

plot(preds)
specs1 <- terra::extract(preds, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

# flag records for manual review. 
specs1 <- split(specs1, f = specs1$scntfcn)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.025, 0.975))
specs2 <- bind_rows(specs1)

specs2 <- specs2 %>% 
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "") 

# compare to see if records already checked for bimodality. 
second <- read.csv('../data/processed/second_phenology_review-scored.csv')
specs3 <- filter(specs2, !accessr %in% second$accessr)
write.csv(specs3, '../data/processed/low_priority_sheets.csv', row.names = F, )

```

### Re-check density curves after manual cleaning again.

Another iterative check. If \>50% of records for some species are still outside of anthesis, we will grab the first/last 10% of records, remove those which have been scored (either \~2.5% or \~5%), and score a final time.

```{r write out training data, eval = F}

scored <- bind_rows(
  read.csv('../data/processed/high_priority_sheets-scored.csv'),
  read.csv('../data/processed/second_phenology_review-scored.csv'), 
  read.csv('../data/processed/low_priority_sheets-scored.csv')
  ) %>% 
  mutate(
    comments = na_if(comments, ""), 
    Anthesis = if_else(is.na(comments), Anthesis, 999), 
    Anthesis = if_else(is.na(Anthesis), 999, Anthesis)
    ) %>% 
  select(-comments)

d1 <- scored %>% # keep records which weren't wonky. 
  group_by(scntfcnm, Anthesis) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs)

dirty <- filter(d1, n() == 2) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)

specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp',
                  quiet = T) %>% 
  filter(scntfcnm %in% unique(dirty$scntfcnm))
pca1 <- terra::rast('../results/spatial/gddPCA.tif')
specs1 <- terra::extract(pca1, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

dirty1 <- filter(scored, scntfcnm %in% dirty$scntfcnm) %>% 
  group_by(scntfcnm) %>% 
  mutate(
    avg_fl = mean(doy), 
    period = if_else(doy < avg_fl, 'E', 'L')
    ) %>% 
  group_by(scntfcnm, Anthesis, period) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs
  ) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)

specs1 <- split(specs1, f = specs1$scntfcnm)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.01, 0.9)) %>% 
  bind_rows() %>% 
  drop_na(phen_flag) %>% 
  filter(! accessr %in% scored$accessr) %>% 
  select(scntfcnm, accessr, doy) %>% 
  st_drop_geometry() %>% 
  arrange(scntfcnm, doy)

write.csv(specs1, '../data/processed/troublesomeSpecies.csv', row.names = F)

rm(scored)
```

### Subset Final training set and create flowering 'absences'

Now create flowering 'absences' for these species. We will start with species which were not flagged via our bimodal detection.

```{r eval= F}

scored <- bind_rows(
  read.csv('../data/processed/high_priority_sheets-scored.csv'),
  read.csv('../data/processed/second_phenology_review-scored.csv'), 
  read.csv('../data/processed/low_priority_sheets-scored.csv'), 
  read.csv('../data/processed/troublesomeSpecies-scored.csv')
  ) %>% 
  mutate(
    Anthesis = if_else(is.na(Anthesis), 999, Anthesis)
    ) %>% 
  select(-comments)

specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  filter(! scntfcnm %in% scored[scored$Anthesis %in% c(0, 999), 'accessr'] )

pca1 <- terra::rast('../results/spatial/gddPCA.tif')
specimens <- terra::extract(pca1, specimens, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

# only re-run the generation of initiation cessation 
# tables for species which were included in the 'low priority'
# and 'troublesome tables

rerun <- bind_rows(
  read.csv('../data/processed/low_priority_sheets-scored.csv'),
  read.csv('../data/processed/troublesomeSpecies-scored.csv')
) %>% 
  distinct(scntfcnm) %>% 
  pull(scntfcnm)

specimens <- filter(specimens, scntfcnm %in% rerun)
splicies <- split(specimens, f = specimens$scntfcnm)
no_cores <- parallel::detectCores()
# write out initiation cessation tables for many species
names(splicies)
lapply(splicies[11:13], ince_writer, bs = 20) 

splicies <- split(specimens, f = specimens$scntfcnm)
no_cores <- parallel::detectCores()
# write out initiation cessation tables for many species
names(splicies)
lapply(splicies[11:13], ince_writer, bs = 20) 
```

```{r catch missed species, eval = F}

p <- '../results/initation_cessation_tables'
tables <- lapply(file.path(p, list.files(path = p)), read.csv)
names(tables) <- gsub('.csv', '', list.files(p))
tables <- data.table::rbindlist(tables, idcol = 'scntfcnm')

specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  bind_rows() %>% 
  filter(! accessr %in% read.csv('../data/processed/manual_review-after2nddensitycurves.csv')$accessr)

# now let's also remove these

processed <- gsub('[.]csv', '', gsub('_', ' ', list.files(p)))
specimens <- #filter(specimens, scntfcnm %in% needed) %>% 
  filter(specimens, ! scntfcnm %in% processed)
pca1 <- terra::rast('../results/spatial/gddPCA.tif')
specimens <- terra::extract(
  terra::rast('../results/spatial/gddPCA.tif'), specimens, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

# write out and review these again based on the second round of  density curves

splicies <- split(specimens, f = specimens$scntfcnm)
no_cores <- parallel::detectCores()

splicies <- Filter(function(x) nrow(x) >= 10, splicies)
names(splicies) # remove ones which are in the folder...
lapply(splicies[2:4], ince_writer, bs = 20) # 41, 40? 42
```

Generate Phenology absences for start and ends of GAM curve for records.

```{r, eval = F}

p <- '../results/initation_cessation_tables'
tables <- lapply(file.path(p, list.files(path = p)), read.csv)
names(tables) <- gsub('[.]csv', '', basename(list.files(path = p)))
tables <- data.table::rbindlist(tables, idcol = 'scntfcnm')

f <- gsub('.shp', '', list.files('../results/PresAbs', pattern = 'shp$'))

specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  bind_rows()
pca1 <- terra::rast('../results/spatial/gddPCA.tif')
specimens <- terra::extract(
  terra::rast('../results/spatial/gddPCA.tif'), specimens, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

tables <- filter(tables, gsub('_', ' ', scntfcnm) %in% specimens$scntfcnm)
specimens <- filter(specimens, scntfcnm %in% gsub('_', ' ', tables$scntfcnm))

specimens <- split(specimens, f = specimens$scntfcnm)
lapply(specimens, pheno_abs_writer)
```




#### Perform scoring for lowest priority species
some species were skipped in the above steps...

```{r}

scored <- bind_rows(
  read.csv('../data/processed/high_priority_sheets-scored.csv'),
  read.csv('../data/processed/second_phenology_review-scored.csv'), 
  read.csv('../data/processed/low_priority_sheets-scored.csv'), 
  read.csv('../data/processed/troublesomeSpecies-scored.csv')
  ) %>% 
  mutate(
    Anthesis = if_else(is.na(Anthesis), 999, Anthesis)
    ) %>% 
  select(-comments)

specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  filter(! scntfcnm %in% scored[scored$Anthesis %in% c(0, 999), 'accessr'] )

```

```{r}
modelled <- list.files('/media/steppe/hdd/SeedPhenology/data/processed/timestamps')

last_species <- read.csv('../data/raw/2024sos_species.csv') %>% 
  mutate(across(.cols = everything(.), ~ na_if(., ''))) %>% 
  unite(taxon, genus:infraspecies, na.rm = TRUE) %>% 
  distinct(taxon) %>% 
  filter(! taxon %in% c(modelled))

speci1 <- filter(specimens, scntfcnm %in% gsub('_', ' ', last_species$taxon))
table(speci1$scntfcnm) # these need modelled - we already have the data

last_species <- last_species %>% 
  mutate(taxon = gsub('_', ' ', taxon)) %>% 
  filter( ! taxon  %in% c(speci1$scntfcnm)) %>% 
  arrange(taxon)

specs1 <- lapply(last_species$taxon, specimen_sampler, date = '1981-01-01')

last_species[ lapply(specs1, typeof) == 'character' , ]

specs1 <- specs1[ lapply(specs1, typeof) == 'list' ] |> bind_rows()
specs1 <- mutate(specs1, scntfcn = str_to_sentence(scientificname)) 
specs1 <- split(specs1, f = specs1$scntfcn)
specs1ABC <- lapply(specs1[1:24], rec_selec)
specs1DEF <- lapply(specs1[25:26], rec_selec)
specs1EFG <- lapply(specs1[28:80], rec_selec)

specs1 <- c(specs1ABC, specs1DEF, specs1EFG)
specs1 <- bind_rows(specs1)

dir.create('../data/processed/lowest_priority_sheets')
sf::st_write(specs1, 
             '../data/processed/lowest_priority_sheets/lowest_priority_sheets.shp', 
             quiet = TRUE, append = F)
```



```{r write out conflicted things for idigbio team}

read.csv('../data/processed/high_priority_sheets-scored.csv')

scored <- bind_rows(
  read.csv('../data/processed/high_priority_sheets-scored.csv'),
  read.csv('../data/processed/second_phenology_review-scored.csv'), 
  read.csv('../data/processed/low_priority_sheets-scored.csv'), 
  read.csv('../data/processed/troublesomeSpecies-scored.csv')
  ) %>% 
  filter(str_detect(accessr, 'prc-symbiota')) %>% 
  select(-Anthesis, -comments) %>% 
  arrange(scntfcnm)

write.csv(scored, '../data/raw/prc_symbiota.csv')
```

