---
title: "Species to model Phenology"
author: "steppe"
date: "2024-01-10"
output: html_document
---

```{r}
library(tidyverse)
library(sf)
source('functions.R')
```

```{r import species}

spp <- read.csv('../data/raw/2024species-Priorities.csv') |>
  unite('scientific_name',  genus:infraspecies, na.rm = TRUE, sep = ' ', remove = F) |>
  mutate(
    scientific_name = trimws(scientific_name)) |>
  select(scientific_name, genus, species, contract, priority, priority_source) |>
  count(scientific_name, priority)

```

```{r pull sheets for species and subset to original, eval = F}

specs1 <- lapply(spp$scientific_name, specimen_sampler, 
                 date = '1981-01-01')

specs1 <- specs1[ lapply(specs1, typeof) == 'list' ] |> bind_rows()
# specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp', quiet = TRUE)
specs1 <- mutate(specs1, scntfcn = str_to_sentence(scientificname)) %>% 
  filter(scientificname != 'Chaenactis artemisifolia')
specs1 <- split(specs1, f = specs1$scntfcn)
specs1 <- lapply(specs1, rec_selec)
specs1 <- bind_rows(specs1)

dir.create('../data/processed/high_priority_sheets')
sf::st_write(specs1, '../data/processed/high_priority_sheets/high_priority_sheets.shp', 
             quiet = TRUE, append = F)

rm(specs1)
```

```{r flag records at outer quantiles of flowering, eval = F}

# extract pca values
preds <- rast('../results/spatial/gddPCA.tif')
specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp')

specs1 <- terra::extract(preds, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

# flag records for manual review. 
specs1 <- split(specs1, f = specs1$scntfcn)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.025, 0.975))
specs2 <- bind_rows(specs1)

priority_spp <- spp %>% 
  filter(priority %in% c('Medium', 'High') | n > 1) %>% 
  pull(scientific_name)

specs2 %>%
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% 
  filter(scntfcnm %in% priority_spp) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "") #%>% 
 # write.csv( '../data/processed/high_priority_sheets.csv', row.names = F, )

rm(preds, specs2, specs1)

as.Date(206, origin = '1987-01-01')
```

the above records were flagged by visual review of the specimen. During this process, the realization that the check for bimodality should have been performed, and the verification that certain species have many more records outside anthesis, called for a second batch of scoring. 

```{r detect multimodal data, eval = F}

d <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp') %>% 
  mutate(scntfcnm = str_to_sentence(scntfcnm))
splicies <- split(d, f = d$scntfcnm)
ou <- sapply(X = splicies, FUN = function(x){diptest::dip(x$doy) }) # this test seems useless. 
ou <- ou[ou < 0.03]
sort(ou, decreasing = F)

ob <- lapply(splicies, modal_finder)
ob <- data.frame(scntfcnm = names(ob), doy = as.numeric(ob))
ob1 <- mutate(ob, doy = case_when(
  scntfcnm == 'Acamptopappus sphaerocephalus' ~ 215,
  scntfcnm == 'Achnatherum speciosum' ~ 250,
  scntfcnm == 'Adenostoma sparsifolium' ~ 999,
  scntfcnm == 'Astragalus eremiticus' ~ 240,
  scntfcnm == 'Atriplex canescens' ~ 999,
  scntfcnm == 'Atriplex confertifolia' ~ 999,
  scntfcnm == 'Balsamorhiza serrata' ~ 180,
  scntfcnm == 'Centromadia elegans' ~ 999,
  scntfcnm == 'Cylindropuntia bigelovii' ~ 210,
  scntfcnm == 'Cylindropuntia bigelovii' ~ 210,
  scntfcnm == 'Danthonia californica' ~ 999,
  scntfcnm == 'Erigeron breweri' ~ 999,
  scntfcnm == 'Lomatium nudicaule' ~ 999,
  scntfcnm == 'Machaeranthera canescens' ~ 999,
  scntfcnm == 'Machaeranthera gracilis' ~ 190,
  scntfcnm == 'Muhlenbegia rigens' ~ 999,
  scntfcnm == 'Penstemon acuminatus' ~ 999,
  scntfcnm == 'Phlox stansburyi' ~ 210,
  scntfcnm == 'Psorothamnus schottii' ~ 225,
  scntfcnm == 'Rosa californica' ~ 999,
  scntfcnm == 'Rubus ursinus' ~ 999,
  scntfcnm == 'Sambucus mexicana' ~ 999,
  scntfcnm == 'Sphaeralcea munroana' ~ 999,
  scntfcnm == 'Stanleya pinnata' ~ 999,
  scntfcnm == 'Symphyotrichum subulatum' ~ 999,
  scntfcnm == 'Triteleia hyacinthina' ~ 999,
  .default = as.numeric(doy)
)) %>% 
  filter(doy != 999)


rm(ou, splicies, ob)
```


```{r identify collections commonly made outside of anthesis, eval = F}

d <- read.csv('../data/processed/high_priority_sheets-scored.csv') %>% 
  mutate(comments = na_if(comments, '')) %>% 
  filter((comments != "str") %>% replace_na(TRUE)) %>% 
  drop_na(Anthesis) 

d1 <- d %>% # keep records which weren't wonky. 
  group_by(scntfcnm, Anthesis) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs)

filter(d1, n() == 1) # totally clean
filter(d1, n() == 2) %>% 
  filter(Anthesis == 0, nAnt <= 0.5)
dirty <- filter(d1, n() == 2) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)

rm(d1)
# for the dirty species, we will determine whether both sides have specimens
# outside the desired phenophase or whether only  a single side does. 
# we will only evaluate specimens on the afflicted end(s). 

dirty1 <- filter(d, scntfcnm %in% dirty$scntfcnm) %>% 
  group_by(scntfcnm) %>% 
  mutate(
    avg_fl = mean(doy), 
    period = if_else(doy < avg_fl, 'E', 'L')
    ) %>% 
  group_by(scntfcnm, Anthesis, period) %>% 
  summarise(nAnt = n()) %>% 
  mutate(
    recs = sum(nAnt),
    nAnt = nAnt / recs
  ) %>% 
  filter(Anthesis == 0, nAnt >= 0.5)
  
rm(dirty, d)
```

Identify records in quantiles. 

```{r identify records for second batch of scoring, eval = F}

# extract pca values
preds <- terra::rast('../results/spatial/gddPCA.tif')
specs1 <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp',
                  quiet = T) %>% 
  filter(scntfcnm %in% c(dirty1$scntfcnm, ob1$scntfcnm))

specs1 <- terra::extract(preds, specs1, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

rm(preds)
# split records with bimodal flowering. 
bimodals <- inner_join(specs1, rename(ob1, 'split' = doy), by = 'scntfcnm') %>% 
  mutate(split = if_else(doy < split, 'EARLY', 'LATE'))

split_bimodals <- split(bimodals, f = ~ bimodals$scntfcnm + bimodals$split)
split_bimodals <- lapply(split_bimodals, visual_review_flagger, probs = c(0.025, 0.975))

sb <- split_bimodals %>%
  map(mutate, phen_flag = as.character(phen_flag)) %>% 
  bind_rows() %>% 
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% # remove records which we already scored, and restrict to 
  # our species of interest.
  filter(scntfcnm %in% ob1$scntfcnm, !accessr %in% d$accessr) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "")

# flag records for further manual review. 
specs1 <- split(specs1, f = specs1$scntfcnm)
specs1 <- lapply(specs1, visual_review_flagger, probs = c(0.05, 0.95))

dirty1 <- ungroup(dirty1)
specs2 <- specs1 %>%
  bind_rows() %>% 
  drop_na(phen_flag) %>% 
  sf::st_drop_geometry() %>% # remove records which we already scored, and restrict to 
  # our speices of interest.
  filter(scntfcnm %in% dirty1$scntfcnm, !accessr %in% d$accessr) %>% 
  select(scntfcnm, accessr, doy) %>% 
  arrange(scntfcnm, doy) %>% 
  mutate(Anthesis = "", # select only the time troublesome time period
         period = if_else(doy < mean(doy), 'E', 'L')) %>% 
  inner_join(., select(dirty1, scntfcnm, period), by = join_by(scntfcnm, period)) %>% 
  select(-period) %>% 
  filter(! accessr %in% sb$accessr)

bind_rows(sb, specs2) %>% 
  arrange(scntfcnm, doy) %>% 
  write.csv(., '../data/processed/second_phenology_review.csv', row.names = F)

rm(sb, specs2, dirty1, split_bimodals, bimodals, ob1, specs1)
```

### Subset final training data set

Based on review of records above, we can now subset the final training set. 

```{r write out training data, eval = F}

scored <- bind_rows(
  read.csv('../data/processed/high_priority_sheets-scored.csv'),
  read.csv('../data/processed/second_phenology_review-scored.csv')
  ) %>% 
  mutate(
    comments = na_if(comments, ""), 
    Anthesis = if_else(is.na(comments), Anthesis, 999), 
    Anthesis = if_else(is.na(Anthesis), 999, Anthesis)
    ) %>% 
  select(-comments)

specimens <- st_read('../data/processed/high_priority_sheets/high_priority_sheets.shp',
                  quiet = T)  %>% 
  select(-scntfcn)

scored <- inner_join(scored, specimens, by = c('scntfcnm', 'doy',  'accessr'))
specimens <- bind_rows(
  filter(specimens, ! accessr %in% scored$accessr) , 
  filter(scored, Anthesis == 1)
  # records which were un-scored are simply dropped - they are neither at Anthesis or not. 
  # further the 'absence' of flowering are dropped, these will come from Weibull. 
) %>% 
  mutate(Anthesis = replace_na(1), 
         scntfcnm = str_to_sentence(scntfcnm))

st_write(specimens, '../data/processed/training/presences.shp', quiet = TRUE)

rm(scored)
```

Now create flowering 'absences' for these species. We will start with species which were not flagged
via our bimodal detection. 

```{r}

trouble <- read.csv('../data/processed/second_phenology_review-scored.csv') %>% 
  pull(scntfcnm) %>% 
  unique()

specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  filter(! scntfcnm %in% trouble)
pca1 <- terra::rast('../results/spatial/gddPCA.tif')

specimens <- terra::extract(pca1, specimens, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

splicies <- split(specimens, f = specimens$scntfcnm)
no_cores <- parallel::detectCores()
# write out initiation cessation tables for many species

splicies <- Filter(function(x) nrow(x) > 300, splicies)
names(splicies)

lapply(splicies[20], ince_writer) 
```


Generate Phenology absences for start and ends of GAM curve for records. 

```{r}

p <- '../results/initation_cessation_tables'
tables <- lapply(file.path(p, list.files(path = p)), read.csv)
names(tables) <- gsub('.csv', '', list.files(p))
tables <- data.table::rbindlist(tables, idcol = 'scntfcnm')


specimens <- st_read('../data/processed/training/presences.shp', quiet = TRUE ) %>% 
  filter(scntfcnm %in% gsub('_', ' ', names(tables)))
pca1 <- terra::rast('../results/spatial/gddPCA.tif')
specimens <- terra::extract(terra::rast('../results/spatial/gddPCA.tif'), specimens, bind = TRUE) |>
  st_as_sf() %>% 
  drop_na(PC1)

specimens <- split(specimens, f = specimens$scntfcnm)

lapply(specimens, pheno_abs_writer)
```




